String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-04-25 2023-1-19/sections/introduction.tex
abstract

\documentclass[11pt]{standalone}
\begin{document}

@@

% the state-of-art string solvers (\cite{z3seq}\cite{z3str3}\cite{z3str4}\cite{z3str3re}\cite{cvc5}\cite{ostrich}\cite{z3trau}\cite{trau}) lose efficiency. 

@@

Z3Str3\cite{z3str3} is a string solver written on the famous SMT solver Z3\cite{z3}. It is based on the DPLL(T) structure and uses the same approach as CVC5, except that the derivation rules of Z3 differ from CVC5. Furthermore, Z3str3 exposes the structure of theory literals to the DPLL(T) SAT solver's branching heuristic, resulting in smarter decisions during its search. Z3Str3 can not solve string constraints (\ref{eq:cvc5_timeout}) as well because the optimization can not avoid the endless unwind of the regular expression. To handle the exponential search space of regular expression, automata-based methods were proposed. One of the automata-based string solver Ostrich\cite{ostrich}\cite{atva2020} encodes string constraints and regular expressions into automata and use the automata theory to solve them. It is sound and complete. But it is inefficient because it is hard to reduce the size of automata. When large counting is constrained on a complex sub-regex, like $ \backslash w \{1, 63\}$ where $\backslash w$ is an English character or a number character, the corresponding automaton in Ostrich becomes very large and can not be solved within 60s time limit. The reason is that Ostrich syntactically rewrites the counting operator to the union of finite repeatings of the sub-regex. For example, $\backslash w\{1, 63\}$ is rewritten to $\backslash w\mid\backslash w^2\cdots\mid\backslash w^{63}$ where $\backslash w^i$ means repeat $i\-th$ times on $\backslash w$. The automaton of $\backslash w$ contains $62$ states to accept each possible character, and the automaton of $\backslash w ^i$ contains $62*i$ states. The total number of states in the automaton is $62 + 62*2 + \cdots 62*63=124992$, causing the automaton to be unsolvable in practice. Trau\cite{z3trau}\cite{trau+} is another powerful automata-based solver. It relies on a Counter-Example Guided Abstraction Refinement (CEGAR) framework in which the under- and over-approximation of the automata are iteratively refined. Although it uses simple flat automata to represent regular expressions, it has to build unsolvable automata on regex $\backslash w\{1, 63\}$ as Ostrich because the solution of the flat automata should be within the solution of the original automata. 
Recently, Z3str3 has been extended to Z3str3RE\cite{z3str3re} by Berzish et al to handle string constants with regular memberships and length operation effectively. It uses automata to formalize the sematic of regular expression and applies several length-aware heuristics. Z3str3RE can avoid building unsolvable automaton because the length abstraction is directly derived from the regex, not the automaton. Unfortunately, the length abstraction is an over-approximation, so it can not solve string constraints (\ref{eq:z3str3re_timeout}) whose regular membership predicate contains complementary.
\begin{equation}\label{eq:z3str3re_timeout}

@@

\subsubsection{Related Work}
To handle the problems in software verification reasoning about strings and integers, string theories with regular membership predicate and linear length constraints are raised. String solvers such as Z3Str3\cite{z3str3}, Z3Seq\cite{z3seq}, Z3Str3RE\cite{z3str3re}, CVC4\cite{cvc4}, CVC5\cite{cvc5}, (Z3-)Trau\cite{trau}\cite{z3trau}, Ostrich(+)\cite{ostrich}\cite{atva2020}, Sloth\cite{sloth}, S3\cite{s3}, and ABC\cite{abc} are developed on these theories. Although string theory is easily
undecidable\cite{undecidable_1}\cite{undecidable_2}, the satisfiability problems for string constraints of regular expressions, linear integer arithmetic, and the string length is decidable\cite{theory_z3str3re}. There are mainly two strategies for solving string constraints: (1) DPLL(T)-based\cite{dpll_t} string solvers\cite{z3str3}\cite{cvc5} apply heuristic derivation rules to unwind the regular expression gradually and identify several classes of simplification techniques for efficiency. (2) Automata-based string solvers\cite{trau}\cite{z3trau}\cite{ostrich}\cite{atva2020}\cite{z3str3re} construct a finite automaton to model the regular expression and use the emptiness problem of the automaton (under linear integer arithmetic) to decide the satisfiability of string constraints.

@@

\end{document}

>>>>

%!TEX root = ../main.tex
%\documentclass[11pt]{standalone}
%\begin{document}

In modern programming languages---such as JavaScript, Python, Java, and PHP---the string data type plays a crucial role. 
It is well-known that string manipulations are error-prone and could even give rise to security vulnerabilities (e.g.\ cross-site scripting, aka XSS). 
One powerful method for identifying such bugs in programs is \emph{symbolic execution} (possibly in combination with dynamic analysis), which analyses symbolic paths in a program by viewing them as constraints, whose feasibility is checked by constraint solvers. 
%
Symbolic execution of string manipulating programs have motivated the highly active research area of \emph{string constraint solving}, resulting in the development of numerous string solvers in the last decade or so including the state-of-the-art string constraint solvers (e.g.
Z3~\cite{Z3}, CVC4/5~\cite{cvc4}, Z3str/2/3/4~\cite{Z3-str,Z3-str2,Z3-str3,BerzishMurphy2021}, Z3str3RE~\cite{BD+23}, 
Z3-Trau~\cite{Z3-trau}, OSTRICH~\cite{CHL+19}, Slent~\cite{WC+18}, among many others.

Regular expressions (abbreviated as regex) and string-length function are widely used in string manipulating programs. According to the statistics from \cite{CS16,DCSL18,WS18}, regular expressions are used in about 30–40\% Java, JavaScript, and Python software projects. 
Moreover, the statistics from \cite{malware_detection_3_kudzu} show that string-length function occupies 78\% of the occurrences of string operations in 18 Javascript applications. 
%Operations like \verb|length| that take string inputs and return an integer frequently appear in real-world JavaScript applications (78\% of string operations in 18 applications)
As a result, most of the aforementioned string constraint solvers support both regular expressions and string-length function. Moreover, specialized algorithms have been proposed to solve efficiently the class of string constraints with regular expressions and string-length function \cite{LTR+15,BD+23}. 
%a decision procedure based on tableaux calculus was designed  in \cite{LTR+15} for the class of string constraints with regular expressions and string-length function and the decision procedure has been implemented in CVC4/5 \cite{LRT+16}.

Counting (aka repetition) is a convenient feature in regular expressions that count the number of matchings of sub-patterns. For instance, $a^{\{2, 4\}}$ specifies that $a$ occurs for at least twice and at most four times and $a^{\{2, \infty\}}$ specifies that $a$ occurs at least twice. 
Note that Kleene star and Kleene plus operator can be seen as special cases of counting operator, for instance, $a^*$ and $a^+$ are equivalent to $a^{\{0,\infty\}}$ and $a^{\{1,\infty\}}$.
Counting is also a frequently used feature of regular expressions. According to the statistics from \cite{CS16}, Kleene star/plus occur in over 70\% of the Python projects, while the other forms of counting occur in over 20\% of them.

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from the string constraints with regex-counting and string-length, especially when the counting bounds are large. For instance, none of CVC4/5, Z3, Z3str3RE, OSTRICH is capable of solving the following constraint within 60 seconds,
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{5, 40\}} (\Sigma \setminus b)^{\{5, 40\}} (\Sigma \setminus c)^{\{5, 40\}} \wedge x \in \Sigma^* c \wedge |x| > 80.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three subwords $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with $c$, and the length of $x$ is greater than $80$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 80$ and the counting upper bound $40$ in  both $(\Sigma \setminus a)^{\{5, 40\}}$ and $(\Sigma \setminus b)^{\{5, 40\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c$ requires that $x$ has to end with $c$.

%counting is used in about 6 percent of the regular expressions in Javascript programs.

\paragraph*{Contribution} xxx

\paragraph*{Related work} xx

\paragraph*{Organization} xxx

\zhilin{stopped here}

@@

% the state-of-art string solvers (\cite{z3seq}\cite{z3str3}\cite{z3str4}\cite{BD+23}\cite{cvc5}\cite{ostrich}\cite{z3trau}\cite{trau}) lose efficiency. 

@@

Z3Str3\cite{z3str3} is a string solver written on the famous SMT solver Z3\cite{Z3}. It is based on the DPLL(T) structure and uses the same approach as CVC5, except that the derivation rules of Z3 differ from CVC5. Furthermore, Z3str3 exposes the structure of theory literals to the DPLL(T) SAT solver's branching heuristic, resulting in smarter decisions during its search. Z3Str3 can not solve string constraints (\ref{eq:cvc5_timeout}) as well because the optimization can not avoid the endless unwind of the regular expression. To handle the exponential search space of regular expression, automata-based methods were proposed. One of the automata-based string solver Ostrich\cite{ostrich}\cite{atva2020} encodes string constraints and regular expressions into automata and use the automata theory to solve them. It is sound and complete. But it is inefficient because it is hard to reduce the size of automata. When large counting is constrained on a complex sub-regex, like $ \backslash w \{1, 63\}$ where $\backslash w$ is an English character or a number character, the corresponding automaton in Ostrich becomes very large and can not be solved within 60s time limit. The reason is that Ostrich syntactically rewrites the counting operator to the union of finite repeatings of the sub-regex. For example, $\backslash w\{1, 63\}$ is rewritten to $\backslash w\mid\backslash w^2\cdots\mid\backslash w^{63}$ where $\backslash w^i$ means repeat $i\-th$ times on $\backslash w$. The automaton of $\backslash w$ contains $62$ states to accept each possible character, and the automaton of $\backslash w ^i$ contains $62*i$ states. The total number of states in the automaton is $62 + 62*2 + \cdots 62*63=124992$, causing the automaton to be unsolvable in practice. Trau\cite{z3trau}\cite{trau+} is another powerful automata-based solver. It relies on a Counter-Example Guided Abstraction Refinement (CEGAR) framework in which the under- and over-approximation of the automata are iteratively refined. Although it uses simple flat automata to represent regular expressions, it has to build unsolvable automata on regex $\backslash w\{1, 63\}$ as Ostrich because the solution of the flat automata should be within the solution of the original automata. 
Recently, Z3str3 has been extended to BD+23\cite{BD+23} by Berzish et al to handle string constants with regular memberships and length operation effectively. It uses automata to formalize the sematic of regular expression and applies several length-aware heuristics. BD+23 can avoid building unsolvable automaton because the length abstraction is directly derived from the regex, not the automaton. Unfortunately, the length abstraction is an over-approximation, so it can not solve string constraints (\ref{eq:BD+23_timeout}) whose regular membership predicate contains complementary.
\begin{equation}\label{eq:BD+23_timeout}

@@

\subsubsection{Related Work}
To handle the problems in software verification reasoning about strings and integers, string theories with regular membership predicate and linear length constraints are raised. String solvers such as Z3Str3\cite{z3str3}, Z3Seq\cite{z3seq}, BD+23\cite{BD+23}, CVC4\cite{cvc4}, CVC5\cite{cvc5}, (Z3-)Trau\cite{trau}\cite{z3trau}, Ostrich(+)\cite{ostrich}\cite{atva2020}, Sloth\cite{sloth}, S3\cite{s3}, and ABC\cite{abc} are developed on these theories. Although string theory is easily
undecidable\cite{undecidable_1}\cite{undecidable_2}, the satisfiability problems for string constraints of regular expressions, linear integer arithmetic, and the string length is decidable\cite{theory_BD+23}. There are mainly two strategies for solving string constraints: (1) DPLL(T)-based\cite{dpll_t} string solvers\cite{z3str3}\cite{cvc5} apply heuristic derivation rules to unwind the regular expression gradually and identify several classes of simplification techniques for efficiency. (2) Automata-based string solvers\cite{trau}\cite{z3trau}\cite{ostrich}\cite{atva2020}\cite{BD+23} construct a finite automaton to model the regular expression and use the emptiness problem of the automaton (under linear integer arithmetic) to decide the satisfiability of string constraints.

@@

%\end{document}

### 2023-04-25 2023-1-19/sections/preliminaries.tex
abstract

\documentclass{standalone}
\begin{document}

@@

\subsubsection{Syntax}
This paper proposes a quantifier-free first-order logic called \textit{Extended String Logic (ESL)}, whose syntax is presented in Table \ref{tab:syntax}. $\varphi$ is a quantifier-free formula that can be a regular membership $x\in\regex$, a quantifier-free Presburger formula $\alpha_1 \leq \alpha_2$, a negation of a formula, or a disjunction of two formulas. $\alpha$ is an integer term which can be an integer constant $m$, integer variable $i$, the length of the string term $|s|$, the minus of an integer term, and the plus of two integer terms. The regular expression $\regex$ is built on empty string $\epsilon$, constant letter $a\in \Sigma$. The supported regex operations involve concatenation $\cdot$, disjunction $+$, intersection $\times$, closure $*$, complement $C$, and bounded repetition $\{m, n\}$. We use $term(\varphi)$ to denote the set of terms and $strvar(\varphi)$ to denote the string variables occurring in $\varphi$. The \emph{literals} of $\varphi$ are $x\in \regex$ and $\alpha_1 \leq \alpha_2$. $x\in \regex$ is called \emph{regular literals} and $\alpha_1 \leq \alpha_2$ is called \emph{linear literals}.
\begin{table}[h]
  \centering
  \begin{tabular}{l r}
    $\varphi$ ::= $x\in \regex \mid \alpha_1\leq\alpha_2 \mid \neg\varphi\mid \varphi\vee\varphi$                                             & formulae            \\
    $\regex$ ::= $\epsilon\mid a\mid \regex\cdot\regex\mid \regex+\regex\mid \regex\times \regex\mid \regex^*\mid\regex^C \mid \regex\{m,n\}$ & regular expressions \\
    $\alpha$ ::= $m \mid i \mid  |s|\mid -\alpha\mid \alpha_1+\alpha_2$                                                                       & integer terms
  \end{tabular}
  \caption{Syntax}\label{tab:syntax}
\end{table}
% Semantic 
\subsubsection{Semantic}
We assume that $S$ is the set of string variables over $\Sigma^*$, and $I$ is the set of integer variables. $\eta: S\times\Sigma\rightarrow\Sigma^*$ is the interpretation on string where $\eta(c)=c$ for every letter $c\in \Sigma$. $\pi: I\rightarrow\mathbb{Z}$ is the interpretation of Presburger arithmetic. Then the semantics is given by a satisfaction relation: $\eta, \pi\models \varphi$ defined in Table \ref{tab:semantics}. We say a formula $\varphi$ is \emph{satisfiable} if a solution $(\eta, \pi)$ exists such as $\eta, \pi\models \varphi$. A formula $\varphi$ is \emph{unsatisfiable} if no solution exists.
\begin{table}[h]
  \centering
  \begin{tabular}{lcl}
    $\eta,\pi \models \varphi_1\vee \varphi_2$ & $\mathsf{iff}$ & $\eta,\pi \models \varphi_1 \text{ or } \eta,\pi \models \varphi_2$ \\
    $\eta,\pi \models \neg\varphi $            & $\mathsf{iff}$ & $\eta,\pi \not\models \varphi$                                      \\
    $\eta, \pi \models x\in \regex$            & $\mathsf{iff}$ & $\exists w \in \lan(\regex),\eta,\pi \models x = w$                 \\
    $\eta, \pi \models \alpha_1 \leq \alpha_2$ & $\mathsf{iff}$ & $\pi(\alpha_1) \leq \pi(\alpha_2) $                                 \\
  \end{tabular}
  \caption{Semantics}
  \label{tab:semantics}
\end{table}
In addition to the classic regex operators (union, concatenation, closure), we syntactically support intersection, complement, and repetition. As we all know, the classic regular language is closed under intersection and complement \cite{aut_hopcraft}. Furthermore, the operation \emph{repetition} $\regex\{m,n\}$ means repeating regex $\regex$ at least $m$ times and at most $n$ times. It can be syntactically rewritten by concatenation and union:
$\regex\{m,n\} \equiv \regex^m\mid\cdots\mid\regex^n$ where $\regex^k$ defines concatenate $\regex$ $k$ times ($m\leq k\leq n$). So the regular language $\lan(\regex)$ defined in Table \ref{tab:syntax} is semantically equal to the classic regular language.

\end{document}

>>>>

%!TEX root = ../main.tex
%\documentclass{standalone}
%\begin{document}

@@

\subsubsection{Syntax}
This paper proposes a quantifier-free first-order logic called \textit{Extended String Logic (ESL)}, whose syntax is presented in Table \ref{tab:syntax}. $\varphi$ is a quantifier-free formula that can be a regular membership $x\in\regex$, a quantifier-free Presburger formula $\alpha_1 \leq \alpha_2$, a negation of a formula, or a disjunction of two formulas. $\alpha$ is an integer term which can be an integer constant $m$, integer variable $i$, the length of the string term $|s|$, the minus of an integer term, and the plus of two integer terms. The regular expression $\regex$ is built on empty string $\epsilon$, constant letter $a\in \Sigma$. The supported regex operations involve concatenation $\cdot$, disjunction $+$, intersection $\times$, closure $*$, complement $C$, and bounded repetition $\{m, n\}$. We use $term(\varphi)$ to denote the set of terms and $strvar(\varphi)$ to denote the string variables occurring in $\varphi$. The \emph{literals} of $\varphi$ are $x\in \regex$ and $\alpha_1 \leq \alpha_2$. $x\in \regex$ is called \emph{regular literals} and $\alpha_1 \leq \alpha_2$ is called \emph{linear literals}.
\begin{table}[h]
  \centering
  \begin{tabular}{l r}
    $\varphi$ ::= $x\in \regex \mid \alpha_1\leq\alpha_2 \mid \neg\varphi\mid \varphi\vee\varphi$                                             & formulae            \\
    $\regex$ ::= $\epsilon \mid a \mid \regex\cdot\regex \mid \regex+\regex \mid \regex^* \mid \regex \cap \regex \mid \regex^C \mid \regex \setminus \regex \mid \regex^{\{m,n\}}$ & regular expressions \\
    $\alpha$ ::= $m \mid i \mid  |s|\mid -\alpha\mid \alpha_1+\alpha_2$                                                                       & integer terms
  \end{tabular}
  \caption{Syntax}\label{tab:syntax}
\end{table}
% Semantic 
\subsubsection{Semantic}
We assume that $S$ is the set of string variables over $\Sigma^*$, and $I$ is the set of integer variables. $\eta: S\times\Sigma\rightarrow\Sigma^*$ is the interpretation on string where $\eta(c)=c$ for every letter $c\in \Sigma$. $\pi: I\rightarrow\mathbb{Z}$ is the interpretation of Presburger arithmetic. Then the semantics is given by a satisfaction relation: $\eta, \pi\models \varphi$ defined in Table \ref{tab:semantics}. We say a formula $\varphi$ is \emph{satisfiable} if a solution $(\eta, \pi)$ exists such as $\eta, \pi\models \varphi$. A formula $\varphi$ is \emph{unsatisfiable} if no solution exists.
\begin{table}[h]
  \centering
  \begin{tabular}{lcl}
    $\eta,\pi \models \varphi_1\vee \varphi_2$ & $\mathsf{iff}$ & $\eta,\pi \models \varphi_1 \text{ or } \eta,\pi \models \varphi_2$ \\
    $\eta,\pi \models \neg\varphi $            & $\mathsf{iff}$ & $\eta,\pi \not\models \varphi$                                      \\
    $\eta, \pi \models x\in \regex$            & $\mathsf{iff}$ & $\exists w \in \lan(\regex),\eta,\pi \models x = w$                 \\
    $\eta, \pi \models \alpha_1 \leq \alpha_2$ & $\mathsf{iff}$ & $\pi(\alpha_1) \leq \pi(\alpha_2) $                                 \\
  \end{tabular}
  \caption{Semantics}
  \label{tab:semantics}
\end{table}
In addition to the classic regex operators (union, concatenation, closure), we syntactically support intersection, complement, and repetition. As we all know, the classic regular language is closed under intersection and complement \cite{aut_hopcraft}. Furthermore, the operation \emph{repetition} $\regex\{m,n\}$ means repeating regex $\regex$ at least $m$ times and at most $n$ times. It can be syntactically rewritten by concatenation and union:
$\regex\{m,n\} \equiv \regex^m\mid\cdots\mid\regex^n$ where $\regex^k$ defines concatenate $\regex$ $k$ times ($m\leq k\leq n$). So the regular language $\lan(\regex)$ defined in Table \ref{tab:syntax} is semantically equal to the classic regular language.

%\end{document}
